---
title: "03. Initial Mapping (Static)"
author: "Caroline Davidson"
format: html
---

# Step 3: Initial Mapping (Static, i.e. Noninteractive)

```{r setup}
#| echo: false
#| include: false
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) # set main project folder as working directory

rm(list=ls())
library(tidyverse)
library(sf)
```

The below code chunk loads the csv datafiles that the previous two scripts create.
```{r load data}

```

# MOVED OVER FROM SCRIPT 2

# Initial Mapping
At this point I'll select the variables of interest in the ACS data and create a subset of the data with only those variables to have a more manageable dataset to work with (there are over 900 variables in the ACS data). 

SNAP specific variables start at index 307 in the metadata file. I selected the following variables for now:

- S2201_C03_001E  /  Estimate: Households receiving food stamps/SNAP: Households
- S2201_C03_021E /  Estimate: Households receiving food stamps/SNAP: Households: POVERTY STATUS IN THE PAST 12 MONTHS: Below poverty level
- S2201_C04_001E  /  Estimate: Percent households receiving food stamps/SNAP: Households
# UPDATE BELOW ONCE I MERGE LINKUP DATA IN!
```{r subset data for relevant SNAP variables}
# COUNTY LEVEL DATA
# creating a subset of my county_acs dataframe to mess around with more easily here:
county_acs_subset <- county_acs %>% 
  select("COUNTY_NAM", "GEO_ID", "NAME", "S2201_C03_001E", "S2201_C03_021E", "S2201_C04_001E",
         "FID", "CO_FIPS", "Shape__Are", "Shape__Len", "geometry")

# all the acs data is characters, resave relevant numbers as numeric
county_acs_subset <- county_acs_subset %>% 
  mutate(S2201_C03_001E = as.numeric(S2201_C03_001E), 
         S2201_C03_021E = as.numeric(S2201_C03_021E), 
         S2201_C04_001E = as.numeric(S2201_C04_001E))

# CENSUS TRACT LEVEL DATA
tract_acs_subset <- tract_acs %>%
  select("GEOID", "COUNTYFP", "Tract_Number", "County", 
         "S2201_C03_001E", "S2201_C03_021E", "S2201_C04_001E",
         "ALAND", "AWATER", "geometry")

# convert relevant ACS data to numeric format
tract_acs_subset <- tract_acs_subset %>%
  mutate(S2201_C03_001E = as.numeric(S2201_C03_001E), 
         S2201_C03_021E = as.numeric(S2201_C03_021E), 
         S2201_C04_001E = as.numeric(S2201_C04_001E))
# 8 NAs introduced for S2201_C04_001E where value was: -
```





```{r initial basic maps - county level}
# map of counties by percent of households receiving food stamps/SNAP
plot(county_acs_subset["S2201_C04_001E"])       


ggplot() +
  geom_sf(data = county_acs_subset,
          aes(fill = S2201_C04_001E, color = S2201_C04_001E)) +
  scale_fill_brewer(palette = "YlGn", name = "Percentage", direction = -1) +
  scale_color_brewer(palette = "YlGn", name = "Percentage", direction = -1) +
  theme_void() +
  labs(
    title = "Percent of Households Receiving Food Stamps / SNAP by County",
    subtitle = "2021 5-Year Estimates from ACS"
  )

```

```{r}
# intermediate materials to send

County_Level_Snap_Data  <- county_acs_subset

County_Level_Snap_Data <- County_Level_Snap_Data %>%
  rename(
    `County Name` = NAME,
    `Number of Households receiving SNAP` = S2201_C03_001E,
    `Number of Households receiving SNAP and below the poverty level` = S2201_C03_021E,
    `Percent of Households receiving SNAP` = S2201_C04_001E
    ) %>%
  select(`County Name`, 
         `Number of Households receiving SNAP`, 
         `Number of Households receiving SNAP and below the poverty level`, 
         `Percent of Households receiving SNAP`)

class(County_Level_Snap_Data) # sf and data.frame, might be easier to select from original df

County_Level_SNAP_Data <- acs_5yr_data_county %>% 
  select(County,
         S2201_C04_001E,
         S2201_C03_001E,
         S2201_C03_021E
         ) %>%
  mutate(
    S2201_C04_001E = as.numeric(S2201_C04_001E),
    S2201_C03_001E = as.numeric(S2201_C03_001E),
    S2201_C03_021E = as.numeric(S2201_C03_021E)
  ) %>%
  arrange(desc(S2201_C04_001E)) %>%
  rename(
    `Number of Households receiving SNAP` = S2201_C03_001E,
    `Number of Households receiving SNAP and below the poverty level` = S2201_C03_021E,
    `Percent of Households receiving SNAP` = S2201_C04_001E
  )

write_csv(County_Level_SNAP_Data, file = "Output/County_Level_SNAP_data.csv")

# plot of counties by percentage of households receiving SNAP benefits
plot(county_acs_subset["S2201_C04_001E"], main = "Percentage of Households receiving SNAP by County") 
```

- S2201_C03_001E  /  Estimate: Households receiving food stamps/SNAP: Households
- S2201_C03_021E /  Estimate: Households receiving food stamps/SNAP: Households: POVERTY STATUS IN THE PAST 12 MONTHS: Below poverty level
- S2201_C04_001E  /  Estimate: Percent households receiving food stamps/SNAP: Households

